Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-21 submission/manuscript.rmd
arxiv style

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
output:
    bookdown::pdf_document2: default
    bookdown::word_document2:
        reference_docx: reference.docx
geometry: margin=1.0in
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{35mm}

Alexandros Rekkas${^1}$, Peter R. Rijnbeek${^1}$, ..., Ewout W. Steyerberg${^2}$, 
David van Klaveren${^3}$

\vspace{40mm}

$^1$ Department of Medical Informatics, Erasmus University Medical Center,
Rotterdam, Netherlands

$^2$ Department of Biomedical Data Sciences, Leiden University Medical Center,
Leiden, Netherlands

$^3$ Department of Public Health, Erasmus University Medical Center, Rotterdam,
Netherlands

\newpage
\newpage
\thispagestyle{empty}
# Abstract {-}
\singlespacing 
**Objective:** Simulation study to compare different risk-based approaches to
estimating individualized treatment effects within the RCT setting. **Study
Design and Setting:** Starting from a base case scenario that assumes a true
constant treatment effect, we considered a total of 66 scenarios for introducing
non-constant effects and evaluating methods under different sample sizes and
baseline risk prediction performance. We compared 7 methods for predicting
absolute benefit: A constant treatment effect model, a risk stratified approach,
a model including a linear interaction of the baseline risk linear predictor
with treatment, 3 restricted cubic spline smoothing models of increasing
flexibility (3, 4 and 5 knots) and an adaptive model selection method based on
Akaike's Information Criterion. We evaluated performance using root mean squared
error, discrimination for benefit and calibration for benefit (i.e., observed
vs. predicted risk difference in treated vs. untreated). **Results:** The model
including a linear interaction of the risk linear predictor with treatment had
adequate performance that was robust under the majority of the simulation
scenarios. Methods using restricted cubic spline smoothing required larger
sample sizes and higher prediction AUC to achieve adequate performance. The
adaptive approach's performance was comparable to the performance of the best
model in each scenario. **Conclusion:** In most cases using a model just
including a linear interaction of the risk linear predictor with treatment
adequately predicts absolute benefit.

\newpage 
\doublespacing

\linenumbers

# Introduction

@@

In the simulated datasets based on the base case scenario treatment was
allocated at random using a 50/50 split. For each patient we simulated $8$
baseline covariates, where $x_1,\dots,x_4\sim N(0, 1)$ and
$x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates.
Coefficient values were such, so that the prediction model had an AUC of $0.75$
and an event rate of $20\%$ in the control arm was achieved. Outcomes in the
treatment arm were created using the same logistic regression model, including a
constant treatment effect odds ratio (OR) of $0.8$. The generated samples of the
base case scenario were of size $n=4,250$ ($80\%$ power for the detection of an
unadjusted OR of $0.8$).

@@

A true logistic regression model with a constant treatment effect (constant OR)
implies that outcome risk in the treatment arm is a straight line parallel to
the first diagonal on the *log-odds* scale, with distance equal to $\log(\text{OR})$. 
(See Figure XX of the supplementary material). We assessed the effect of
stronger and absent relative treatment effects ($\text{OR}=0.5$ or
$\text{OR}=1$). We also relaxed the assumption that the line should be parallel
to the diagonal, considering moderate and stronger linear deviations. Finally,
we dropped the assumption of linearity allowing for quadratic deviations.

We also considered scenarios with treatment-covariate interactions. These
scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

@@

We also compared the predictive performance of the methods under study. We
assessed discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specific *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
considered (Figure \@ref(fig:rmsebase); Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \@ref(fig:rmsebase); Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in each scenario.

@@

When we increased the sample size (N = `r 17000`), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\@ref(fig:rmsesamplesize); Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\@ref(fig:rmsesamplesize); Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \@ref(fig:rmsesamplesize); Panel D).

@@

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\@ref(fig:rmseauc); Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \@ref(fig:rmseauc); Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

@@

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
including a linear interaction with the risk linear predictor tended to present
much lower variability compared to all other model-based and smoothing
approaches (Figure \@ref(fig:discrimination)). We also observed an increasing
trend of discrimination for benefit variability with increasing number of
restricted cubic spline knots in all scenarios.

@@

When focusing on calibration, the linear interaction model had the lowest median
ICI for benefit in the majority of scenarios except for the scenarios where
moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated the best performance, very
comparable to the linear interaction model's performance, nonetheless (Figure
\@ref(fig:calibration); Panels A, B, and C).

>>>>

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare different risk-based approaches to estimating
  individualized treatment effects with simulations of the RCT setting. **Study
  Design and Setting:** We compared different methods for predicting absolute
  benefit on the basis of an available prognostic index (PI) for risk: A model
  with the PI and a constant relative treatment effect; models including an
  interaction of treatment assignment with the PI; 4 quarters of the PI; nonlinear
  transformations of the PI with increasing flexibility (restricted cubic splines
  with 3, 4 and 5 knots, respectively); an adaptive method, selecting the optimal
  model based on Akaike’s Information Criterion. Starting from a base case
  scenario (sample size 4,250, average treatment effect odds ratio 0.8, AUC of the
  PI 0.75), we considered diverse assumptions of the data generation process of
  the “true” treatment effect, introducing linear and quadratic interactions of
  the PI with treatment. We analyzed the effect of varying sample size and
  discriminative ability of the PI. We evaluated performance using root mean
  squared error, discrimination for benefit and calibration for benefit. **Results:**
  The model including a linear interaction of the risk linear predictor with
  treatment had adequate performance that was robust under the majority of the
  simulation scenarios. Methods using restricted cubic splines required larger
  sample sizes and higher AUC of the PI to achieve adequate performance. The
  adaptive approach’s performance was comparable to the performance of the best
  model in each scenario. **Conclusion:** In most cases using a model just
  including a linear interaction of the PI with treatment
  adequately predicts absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - test
  - test that
output: 
  rticles::arxiv_article:
    keep_tex: true
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```

\linenumbers
# Introduction

@@

In the simulated datasets based treatment was allocated at random using a 50/50
split. For each patient we simulated $8$ baseline covariates, where
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes
for patients in the control arm were generated from a logistic regression model
including all baseline covariates. Coefficient values were such, so that the
prediction model had an AUC of $0.75$ and an event rate of $20%$ in the control
arm was achieved. Under the base case scenario, outcomes in the treatment arm
were created using the same logistic regression model, including a constant
treatment odds ratio (OR) of $0.8$. The generated samples of the base case
scenario were 4,250 ($80\%$ power for the detection of an unadjusted OR of
$0.8$).

@@

We simulated binary outcomes in the control arm based on 
$\text{expit}(PI)$, where 
$\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
In the treatment arm outcomes where generated from $\text{expit}(lp_1)$ with
$$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
where the coefficients $\gamma_0, \gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario. In this way we
were able to assess a wide variety of scenarios, ranging from true constant
relative treatment effects to moderate and strong linear and quadratic
deviations. We also considered scenarios with treatment-covariate interactions.
These scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

@@

We also compared the predictive performance of the methods under study. We
assessed discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specifi *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
considered (Figure \ref{fig:rmsebase}; Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \ref{fig:rmsebase}; Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in each scenario.

@@

When we increased the sample size (N = `r 17000`), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \ref{fig:rmsesamplesize}; Panel D).

@@

When we increased the true prediction AUC to 0.85, models including RCS
smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations.

@@

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
including a linear interaction with the risk linear predictor tended to present
much lower variability compared to all other model-based and smoothing
approaches (Figure \ref{fig:discrimination}). We also observed an increasing
trend of discrimination for benefit variability with increasing number of
restricted cubic spline knots in all scenarios.

@@

When focusing on calibration, the linear interaction model had the lowest median
ICI for benefit in the majority of scenarios except for the scenarios where
moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated the best performance, very
comparable to the linear interaction model's performance, nonetheless (Figure
\ref{fig:calibration}; Panels A, B, and C).
